WASP-4 is Accelerating Toward the Earth

### 2020-03-19 paper/ms.tex
update paper to describe better AO constraints

We observed WASP-4 twice, on the night of September 11-12 with
relatively poor seeing (1.2$''$) and also on the night of September
28-29.  On each night, we acquired three sets of $1000\times
60$$\,$msec exposures.  If a companion is present, the autocorrelation
functions of these speckle images would reveal a characteristic
interference pattern. This pattern is then used to determine the
properties of the detected companion and produce a reconstructed
image.  Using the reconstructed speckle images, contrast curves are
produced to determine the 5-$\sigma$ detection limits (see
\citealt{howell_speckle_2011}).  No companions were detected. The data
from the second night, which had better seeing (0.6$''$), led to the
most constraining limit.  The 832$\,$nm limits were the most useful,
given that any faint companion would likely be redder than the host
star.  Therefore, we opted to use the 832$\,$nm September 28-29
contrast limits in the analysis described below.

@@

We fitted each model by assuming a Gaussian likelihood and sampling
over the posterior probability distributions.  We sampled the
posterior using the algorithm proposed by
\citet{goodman_ensemble_2010} and implemented by
\citet{foreman-mackey_emcee_2013} in \texttt{emcee}.  The prior for
the quadratic model allowed the period derivative to be either
positive or negative.

@@

From the high values of $\chi^2_{\rm red}$, we can surmise that
neither model provides a satisfactory fit to the transit data; there
must be some additional signal or noise.  The earlier study by
\citetalias{bouma_wasp4b_2019} found that the quadratic model for the
(sparser) transit data gave $\chi^2_{\rm red}=1.0$.  The worsening of
$\chi^2_{\rm red}$ probably reflects underestimated measurement
uncertainties in at least some of the newly reported transit times,
due to time correlated noise in the photometry. It might also be due
to errors or misunderstandings of the time systems in which the data
were recorded, despite our best efforts to interpret the literature.
In what follows, we opted to enlarge the uncertainties in the
best-fitting parameters of each model by a factor of $(\chi^2_{\rm
red})^{1/2}$.  This was a factor of $\approx$1.73$\times$ for the
linear model, and $\approx$1.41$\times$ for the quadratic model.

@@

@@

\paragraph{Speckle imaging constraints}

First, we converted the contrast curves obtained from speckle imaging
(Figure~\ref{fig:zorro}) to upper limits on the mass of a companion as
a function of separation. To do this, we followed
\citet{montet_trends_2014}, and opted to employ the
\citet{baraffe_evolutionary_2003} models for substellar mass objects
and the MIST isochrones for stellar mass objects
\citep{paxton_modules_2011,paxton_modules_2013,paxton_modules_2015,dotter_mesa_2016,choi_mesa_2016}.
We assumed that the system age was 5 Gyr, at which point the companion
would have fully contracted.

Due to the custom filters of the Zorro imager, and corresponding lack
of synthetic photometry, we further assumed that all sources had
blackbody spectra. While this is a simplification, we do not have
ready access to the planetary and stellar atmosphere models needed for
the consistent calculation with the \texttt{COND03} and \texttt{MESA}
models.  We adopted the effective temperatures and bolometric
luminosities from the \citet{baraffe_evolutionary_2003} and MIST
isochrones.  Using these theoretical quantities and the
empirically-measured Zorro bandpasses, we calculated absolute
magnitudes in the 562 and 832 nm Zorro bands for stellar and planetary
mass companions.  Applying the same calculation to WASP-4 itself using
the effective temperature and bolometric luminosity from
\citetalias{bouma_wasp4b_2019}, we derived the transformation from
contrast ratio to companion mass.  The resulting limits are shown in
Figure~\ref{fig:mass_sma}.

\paragraph{Radial velocity constraints}

To derive constraints on possible companion masses and separations
from the radial velocities, we mostly followed the procedure of
\citet{bryan_excess_2019}. We began by defining a $128\times128$ grid
in true planetary mass and semimajor axis, over a logarithmic grid
ranging from 1 to 900$\,$$M_{\rm Jup}$ and 3 to 500$\,$${\rm AU}$.  We
then considered the possibility that an additional companion in any
particular cell could explain the observed linear trend.

@@

For each simulated companion, we drew a set of model parameters from
the converged posterior probability samples of our model for the
WASP-4 radial velocities.  We subtracted the orbital component of the
model from the observed RVs, leaving behind the RV residuals with a
linear trend.  Given $(a_{\rm c}, M_{\rm c}, e_{\rm c})$ for each
simulated outer companion, and the choice of instrument offsets and
jitters, we performed a maximum likelihood fit for the time and
argument of periastron of the outer simulated companion.  We converted
the resulting $128\times128\times512$ cube of log-likelihood values to
probabilities, and averaged over the samples in each grid cell to
derive a probability distribution in mass and semi-major axis.
Figure~\ref{fig:mass_sma} shows the result\replaced{.}{: the companion
responsible for the acceleration has a true mass of 10-200$\,M_{\rm
Jup}$ and an orbital distance of $10$-$100\,{\rm AU}$.}

@@

The previously offered explanations for WASP-4b's decreasing orbital
period included tidal orbital decay, apsidal precession, and the
Doppler effect \citep{bouma_wasp4b_2019}.  Our new radial velocity
measurements strongly indicate that the least exotic option---the
Doppler effect---is the \replaced{most likely explanation}{dominant
physical process}.  Long-term transit timing data show the orbital
period to be decreasing by $- 8.64 \pm 1.26~{\rm msec}\,{\rm
yr}^{-1}$.  The long-term radial velocity data show a trend that
should lead to an apparent period derivative of $-5.94 \pm 0.39~{\rm
msec}\,{\rm yr}^{-1}$ through the Doppler effect.  Although these two
measurements of the period derivative are discrepant by about
2-$\sigma$, Occam's razor \replaced{suggests that most or all of}{would suggest that} the
apparent decrease of WASP-4b's orbital period is caused
\added{solely} by the line-of-sight acceleration.\added{ Detection
of any additional second-order effects will require more precise data,
and a more significant discrepancy.}

>>>>

We observed WASP-4 twice, on the night of September 11-12 with
relatively poor seeing (1.2$''$) and also on the night of September
28-29\added{, 2019}.  On each night, we acquired three sets of
$1000\times 60$$\,$msec exposures.  If a companion is present, the
autocorrelation functions of these speckle images would reveal a
characteristic interference pattern. This pattern is then used to
determine the properties of the detected companion and produce a
reconstructed image.  Using the reconstructed speckle images, contrast
curves are produced to determine the 5-$\sigma$ detection limits (see
\citealt{howell_speckle_2011}).  No companions were detected. The data
from the second night, which had better seeing (0.6$''$), led to the
most constraining limit.  The 832$\,$nm limits were the most useful,
given that any faint companion would likely be redder than the host
star.  Therefore, we opted to use the 832$\,$nm September 28-29
contrast limits in the analysis described below.

@@

We fitted each model by assuming a Gaussian likelihood and sampling
over the posterior probability distributions.  
\added{The transit timing measurements, uncertainties, and provenances
are given in Table~1.}
We sampled the posterior using the algorithm proposed by
\citet{goodman_ensemble_2010} and implemented by
\citet{foreman-mackey_emcee_2013} in \texttt{emcee}.  The prior for
the quadratic model allowed the period derivative to be either
positive or negative.

@@

From the high values of $\chi^2_{\rm red}$, we can surmise that
neither model provides a satisfactory fit to the transit data; there
must be some additional signal or noise.  The earlier study by
\citetalias{bouma_wasp4b_2019} found that the quadratic model for the
(sparser) transit data gave $\chi^2_{\rm red}=1.0$.  The worsening of
$\chi^2_{\rm red}$ \replaced{probably reflects}{could reflect} underestimated \replaced{measurement}{statistical}
uncertainties in at least some of the newly reported transit times,
\added{perhaps} due to time correlated noise in the photometry. It
might also be due to \replaced{errors or}{systematic} misunderstandings of the time systems
in which the data were recorded, despite our best efforts to interpret
the literature. \added{In particular, mistaken leap-second corrections can introduce systematic errors of order one minute.}  \replaced{In what follows,}{Since we have no way of
identifying which observations are affected by these issues,} we opted to
\added{uniformly} enlarge the
uncertainties in the best-fitting parameters of each model by a factor
of $(\chi^2_{\rm red})^{1/2}$.  This was a factor of
$\approx$1.73$\times$ for the linear model, and $\approx$1.41$\times$
for the quadratic model.

@@

\subsection{Radial velocities}
\label{subsec:rvs}

@@

%\paragraph{Speckle imaging constraints}
% \replaced{\paragraph{Speckle imaging constraints}}{
\paragraph{Speckle imaging transformations}

First, we converted the contrast curves obtained from speckle imaging
(Figure~\ref{fig:zorro}) to upper limits on the mass of a companion as
a function of \added{projected} separation. To do this, we followed
\citet{montet_trends_2014}, and opted to employ the
\citet{baraffe_evolutionary_2003} models for substellar mass objects
and the MIST isochrones for stellar mass objects
\citep{paxton_modules_2011,paxton_modules_2013,paxton_modules_2015,dotter_mesa_2016,choi_mesa_2016}.
We assumed that the system age was 5 Gyr, at which point the companion
would have fully contracted.

Due to the custom filters of the Zorro imager, and corresponding lack
of synthetic photometry, we further assumed that all sources had
blackbody spectra. While this is a simplification, we do not have
ready access to the planetary and stellar atmosphere models needed for
the consistent calculation with the \texttt{COND03} and \texttt{MESA}
models.  We adopted the effective temperatures and bolometric
luminosities from the \citet{baraffe_evolutionary_2003} and MIST
isochrones.  Using these theoretical quantities and the
empirically-measured Zorro bandpasses, we calculated absolute
magnitudes in the 562 and 832 nm Zorro bands for stellar and planetary
mass companions.  Applying the same calculation to WASP-4 itself using
the effective temperature and bolometric luminosity from
\citetalias{bouma_wasp4b_2019}, we derived the transformation from
contrast ratio to companion mass.  The resulting
limits \added{derived if we assume maximal projected separations} are
\replaced{shown}{shown with the blue line} in Figure~\ref{fig:mass_sma}.
\added{We describe our procedure for incorporating projection effects
in the following paragraphs.}

%\replaced{\paragraph{Radial velocity constraints}}
\paragraph{Combined radial velocity and imaging constraints}
To derive constraints on possible companion masses and separations
\deleted{from the radial velocities}, we mostly followed the procedure of
\citet{bryan_excess_2019}. We began by defining a $128\times128$ grid
in true planetary mass and semimajor axis, over a logarithmic grid
ranging from 1 to 900$\,$$M_{\rm Jup}$ and 3 to 500$\,$${\rm AU}$.  We
then considered the possibility that an additional companion in any
particular cell could explain the observed linear trend.

@@

\explain{Modified wording for clarity.}
\replaced{For each simulated companion, we drew a set of model
parameters from the converged posterior probability samples of our
model for the WASP-4 radial velocities.}{The orbital properties of the
inner hot Jupiter were assigned for each simulated system by sampling
from from the radial velocity posterior derived in
Section~\ref{subsec:rvs}.}  We subtracted the orbital component of the
model from the observed RVs, leaving behind the RV residuals with a
linear trend.  Given $(a_{\rm c}, M_{\rm c}, e_{\rm c})$ for each
simulated outer companion, and the choice of instrument offsets and
jitters, we performed a maximum likelihood fit for the time and
argument of periastron of the outer simulated companion.  We converted
the resulting $128\times128\times512$ cube of log-likelihood values to
probabilities, and averaged over the samples in each grid cell to
derive a probability distribution in mass and semi-major axis, based
on the radial velocities.

\added{We incorporated the speckle imaging limits in each simulated
system as follows. After fitting for the time and argument of
periastron, all the orbital parameters needed to find the projected
separation are known (we assumed uniform sensitivity as a function of
position angle, and therefore fixed the longitude of the
ascending node to zero). We calculated the projected separation using
the parametrization given by \citet{quirrenbach_2010}, which required
solving Kepler's equation for each simulated system at the time of the
imaging observation (September 28, 2019).   For each cell, we then
calculated the fraction of companions whose masses and
projected separations would have enabled detection.  }

\replaced{Figure~\ref{fig:mass_sma} shows the result.}
{The contours in Figure~\ref{fig:mass_sma} show the result of multiplying the radial
velocity and speckle imaging probability distributions. The companion
responsible for the acceleration has a true mass of 10-200$\,M_{\rm
Jup}$ and an orbital distance of $10$-$100\,{\rm AU}$.}
%FIXME: might need to update this text!

@@

The previously offered explanations for WASP-4b's decreasing orbital
period included tidal orbital decay, apsidal precession, and the
Doppler effect \citep{bouma_wasp4b_2019}.  Our new radial velocity
measurements strongly indicate that the least exotic option---the
Doppler effect---is the \replaced{most likely explanation}{dominant
physical process}.  Long-term transit timing data show the orbital
period to be decreasing by $- 8.64 \pm 1.26~{\rm msec}\,{\rm
yr}^{-1}$.  The long-term radial velocity data show a trend that
should lead to an apparent period derivative of $-5.94 \pm 0.39~{\rm
msec}\,{\rm yr}^{-1}$ through the Doppler effect.  Although these two
measurements of the period derivative are discrepant by about
2-$\sigma$, Occam's razor \replaced{suggests that most or all of}{would suggest that} the
apparent decrease of WASP-4b's orbital period is caused
\added{solely} by the line-of-sight acceleration.\added{ Detection
of additional second-order effects will require more precise data,
and a more significant discrepancy.}
